---
title: "Determining the causal effect of kidney function on diastolic blood pressure using two sample mendelian randomization approach"
author: " Irene, Olga, Victor, Dr.Allan and Chris"
date: "3/23/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## INTRODUCTION

# Two sample Mendelian randomization using the TwoSampleMR package

## Install and load required packages

```{r}
#install.packages("TwoSampleMR")
library(dplyr)
library(data.table)
library(TwoSampleMR)
library(ggplot2)
library(plyr) 
library(devtools)
library(calibrate)
library(ggrepel)
library(ggthemes)
library(MRInstruments)
library(png)
```

## Set working directory
``` {r}
setwd("/home/irene/MR_UG")
```

### Load in Exposure Data
```{r}
rm(list = ls())
mvpexp <- read.csv("signMVP_header.txt", h=T, sep = "\t")
#mvpexp
dbp_out <- read.table("DBP_greped2.txt",h=T,sep = " ")
for_merge <- read.table("for_merge", h=T)
```

### merge exp n outcome data
```{r, include=TRUE}
merged <- merge(dbp_out,for_merge,by = "snpid")
colnames(mvpexp) <- paste("EXP", colnames(mvpexp), sep = "_")
colnames(dbp_out) <- paste("OUT", colnames(dbp_out), sep = "_")

```


```{r, include=TRUE}
colnames(mvpexp)
colnames(dbp_out)

```
### Format the Exposure data to comply with TSMR package
```{r}
expformated <- format_data(mvpexp, header = TRUE, type = "exposure", snp_col = "EXP_rs", beta_col = "EXP_EffectEstimate", se_col = "EXP_SE", effect_allele_col = "EXP_EA", other_allele_col = "EXP_Allele2", eaf_col = "EXP_EAF", pval_col = "EXP_Pvalue")
```

### Load the Outcome Data
```{r}
# alt_out <- read.table("alt_outcome.txt", h=T,quote="\"", comment.char="")
```
### View the structure of the outcome data using str and head

```{r}
# str(alt_out)
# head(alt_out)
colnames(merged)
```

### Format the Outcome data to comply with TSMR package
```{r}
outformated <- format_data(merged, header = TRUE, type = "outcome", snp_col = "rs", beta_col = "beta_uganda", se_col = "se_uganda", effect_allele_col = "allele1", other_allele_col = "allele2", eaf_col = "af_uganda", pval_col = "pval_uganda")
```

### Harmonize both exposure and outcome
*Note: there are different harmonization fuction depending on what we want to do with palindromic SNPs. This shall be explained in details.*
```{r}
harmonized <- harmonise_data(exposure_dat = expformated, outcome_dat = outformated, action = 2)

#write.table(harmonized, file = "harmonized.txt", row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```
```{r}
# palindromic_at<-subset(harmonized,effect_allele.exposure %in% "A"&other_allele.exposure %in% "T")
# palindromic_ta<-subset(harmonized,effect_allele.exposure %in% "T"&other_allele.exposure %in% "A")
# palindromic_gc<-subset(harmonized,effect_allele.exposure %in% "G"&other_allele.exposure %in% "C")
# palindromic_cg<-subset(harmonized,effect_allele.exposure %in% "C"&other_allele.exposure %in% "G")
# dim(palindromic_at)
# dim(palindromic_ta)
# dim(palindromic_gc)
# dim(palindromic_cg)
```

### Perform clumping
*Also take note of the different clumping threshold. Vary the clumping threshold and see how that impacts the results.*
```{r}
Exp_Out_Harmonized_Clumped <- clump_data(harmonized,clump_kb = 500000,clump_r2 = 0.01,clump_p1 = 1,clump_p2 = 1, pop = "AFR")
```

### Perform MR analysis
```{r}
Exp_Out_Harmonized_Clumped2 <- clump_data(harmonized,clump_kb = 10000,clump_r2 = 0.1,clump_p1 = 1,clump_p2 = 1, pop = "AFR")
```

```{r}
mr_results <- mr(Exp_Out_Harmonized_Clumped, method_list=c("mr_ivw","mr_egger_regression","mr_weighted_median", "mr_weighted_mode"))
mr_results
results <-cbind.data.frame(mr_results$outcome,mr_results$nsnp,mr_results$method,mr_results$b,mr_results$se,mr_results$pval)

mr_results2 <- mr(Exp_Out_Harmonized_Clumped2, method_list=c("mr_ivw","mr_egger_regression","mr_weighted_median", "mr_weighted_mode"))
mr_results2
results2 <-cbind.data.frame(mr_results$outcome,mr_results$nsnp,mr_results$method,mr_results$b,mr_results$se,mr_results$pval)
```

```{r}
write.csv(results,"dbp_kidney.csv")
write.csv(results2,"dbp_kidney2.csv")
```

## Sensitivity analysis
```{r}
#heterogeneity
het <- mr_heterogeneity(Exp_Out_Harmonized_Clumped)
het
#pleiotropy
pleio <- mr_pleiotropy_test(Exp_Out_Harmonized_Clumped)
pleio
res_single <- mr_singlesnp(Exp_Out_Harmonized_Clumped)
res_single

#heterogeneity
het2 <- mr_heterogeneity(Exp_Out_Harmonized_Clumped2)
het2
#pleiotropy
pleio2 <- mr_pleiotropy_test(Exp_Out_Harmonized_Clumped2)
pleio2
res_single2 <- mr_singlesnp(Exp_Out_Harmonized_Clumped2)
res_single2
```
## Generate Plots
```{r, include= TRUE}
png("dbp_kidney.png")
  mr_scatter_plot(mr_results, Exp_Out_Harmonized_Clumped)
dev.off()

png("dbp_kidney2.png")
  mr_scatter_plot(mr_results2, Exp_Out_Harmonized_Clumped2)
dev.off()
```

```{r}
png("dbp_kidney_forest.png")
  mr_forest_plot(res_single)
dev.off()

png("dbp_kidney_forest2.png")
  mr_forest_plot(res_single2)
dev.off()
```

```{r}
png("dbp_kidneyfunnel.png")
  mr_funnel_plot(res_single)
dev.off()

png("dbp_kidneyfunnel2.png")
  mr_funnel_plot(res_single2)
dev.off()

```
```{r}
res_loo <- mr_leaveoneout(Exp_Out_Harmonized_Clumped)
png("dbp_kidney_loo.png")
  mr_leaveoneout_plot(res_loo)
dev.off()

res_loo2 <- mr_leaveoneout(Exp_Out_Harmonized_Clumped2)
png("dbp_kidney_loo2.png")
  mr_leaveoneout_plot(res_loo2)
dev.off()
```
